---
title: "Olympic Running Times Analysis"
author: "Applied Forecasting - Spring 2025"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(fpp3)
```

## Question 1: Olympic Running Times

### 1.1 Plot the winning time against the year for each event (Length + Sex)

```{r}
# Load the dataset
data("olympic_running")

# Create Event variable
data <- olympic_running %>%
  mutate(Event = paste(Length, Sex))

# Plot winning times by event
ggplot(data, aes(x = Year, y = Time)) +
  geom_line() +
  facet_wrap(~Event, scales = "free_y") +
  labs(title = "Winning Times in Olympic Running Events",
       y = "Time (seconds)", x = "Year")
```

### 1.2 Fit a regression line to each event

```{r}
# Fit linear regression model by Event
lm_models <- data %>%
  model(lm_model = TSLM(Time ~ Year))

# Extract slope coefficients
lm_models %>%
  tidy() %>%
  filter(term == "Year")
```

### 1.3 Plot the residuals

```{r}
library(fpp3)

# Reload the dataset from scratch to avoid accidental mutation
data("olympic_running")

# Step 1: Create an Event variable and convert to grouped tsibble
olympic_grouped <- olympic_running %>%
  mutate(Event = paste(Length, Sex)) %>%
  as_tsibble(key = Event, index = Year)

# Step 2: Fit a regression model for each event
lm_models <- olympic_grouped %>%
  model(lm_model = TSLM(Time ~ Year))

# Step 3: Augment + Residual Plot
lm_models %>%
  augment() %>%
  ggplot(aes(x = Year, y = .resid)) +
  geom_line() +
  facet_wrap(~Event, scales = "free_y") +
  labs(title = "Residuals from Linear Regression",
       y = "Residuals", x = "Year")

```

### 1.4 Forecast for 2020 and 2024

```{r}
# Step 1: Get future years and combine with keys
future_olympics <- olympic_grouped %>%
  distinct(Event) %>%
  crossing(Year = c(2020, 2024)) %>%
  as_tsibble(index = Year, key = Event)

# Step 2: Forecast using the model
forecasts <- lm_models %>%
  forecast(new_data = future_olympics)

# Step 3: Plot forecasts
event_list <- unique(olympic_grouped$Event)

for (e in event_list) {
  print(
    forecasts %>%
      filter(Event == e) %>%
      autoplot(olympic_grouped %>% filter(Event == e)) +
      labs(title = paste("Forecast for", e, "(2020 & 2024)"),
           y = "Time (seconds)", x = "Year") +
      theme_minimal()
  )
}


```
## Question 2: Elasticity Coefficient in Log-Log Model

An elasticity coefficient is the ratio of the percentage change in the forecast variable $y$ to the percentage change in the predictor variable $x$. Mathematically, the elasticity is defined as $\frac{dy}{dx} \cdot \frac{x}{y}$.

The question asks us to consider the log-log model: $\log(y_t) = \beta_0 + \beta_1\log(x_t) + \epsilon_t$ and show that $\beta_1$ is the elasticity coefficient.

### Solution:

First, I'll express $y$ as a function of $x$ by working with the log-log model equation:

$\log(y_t) = \beta_0 + \beta_1\log(x_t) + \epsilon_t$

To isolate $y_t$, I need to exponentiate both sides:

$y_t = \exp(\beta_0 + \beta_1\log(x_t) + \epsilon_t)$

Using the properties of exponentials:

$y_t = \exp(\beta_0) \times \exp(\beta_1\log(x_t)) \times \exp(\epsilon_t)$

Since $\exp(\beta_1\log(x_t)) = x_t^{\beta_1}$, I get:

$y_t = \exp(\beta_0) \times x_t^{\beta_1} \times \exp(\epsilon_t)$

Now, to show that $\beta_1$ is the elasticity coefficient, I need to calculate:
$\text{elasticity} = \frac{dy}{dx} \cdot \frac{x}{y}$

First, I'll find $\frac{dy}{dx}$ by differentiating $y$ with respect to $x$:
$\frac{dy}{dx} = \frac{d}{dx}[\exp(\beta_0) \times x_t^{\beta_1} \times \exp(\epsilon_t)]$

Since $\exp(\beta_0)$ and $\exp(\epsilon_t)$ are constants with respect to $x$:
$\frac{dy}{dx} = \exp(\beta_0) \times \exp(\epsilon_t) \times \frac{d}{dx}[x_t^{\beta_1}]$
$\frac{dy}{dx} = \exp(\beta_0) \times \exp(\epsilon_t) \times \beta_1 \times x_t^{(\beta_1-1)}$

Now, I'll calculate the elasticity:
$\text{elasticity} = \frac{dy}{dx} \cdot \frac{x}{y}$
$\text{elasticity} = [\exp(\beta_0) \times \exp(\epsilon_t) \times \beta_1 \times x_t^{(\beta_1-1)}] \times [\frac{x_t}{(\exp(\beta_0) \times x_t^{\beta_1} \times \exp(\epsilon_t))}]$

Simplifying:
$\text{elasticity} = \beta_1 \times x_t^{(\beta_1-1)} \times \frac{x_t}{x_t^{\beta_1}}$
$\text{elasticity} = \beta_1 \times x_t^{(\beta_1-1+1-\beta_1)}$
$\text{elasticity} = \beta_1 \times x_t^0$
$\text{elasticity} = \beta_1$

Therefore, in a log-log model, the coefficient $\beta_1$ represents the elasticity of $y$ with respect to $x$.


## Question 3: US Gasoline Time Series

### 3.1 Fit harmonic regression with trend

```{r}
data("us_gasoline")

gasoline_2004 <- us_gasoline %>%
  filter(Week <= yearweek("2004 W52"))

harmonic_model_k2 <- gasoline_2004 %>%
  model(TSLM(Barrels ~ trend() + fourier(period = 52, K = 2)))

harmonic_model_k2 %>%
  augment() %>%
  ggplot(aes(x = Week)) +
  geom_line(aes(y = Barrels), color = "black") +
  geom_line(aes(y = .fitted), color = "blue") +
  labs(title = "Harmonic Regression (K=2)", y = "Million Barrels per Day", x = "Week")
```

### 3.2 Select best number of Fourier terms by AICc

```{r}
aicc_values <- tibble(K = 1:6) %>%
  mutate(
    model = map(K, ~ {
      gasoline_2004 %>%
        model(TSLM(Barrels ~ trend() + fourier(period = 52, K = .x)))
    }),
    AICc = map_dbl(model, ~ glance(.x)$AICc)
  )

aicc_values
```

### 3.3 Residual diagnostics and Ljung-Box test

```{r}
best_model <- gasoline_2004 %>%
  model(TSLM(Barrels ~ trend() + fourier(period = 52, K = 3)))

best_model %>%
  gg_tsresiduals()

augment(best_model) %>%
  features(.resid, feat = ljung_box, lag = 24, dof = 6)
```

### 3.4 Forecast for 2005

```{r}
forecast_2005 <- best_model %>%
  forecast(h = "1 year")

forecast_2005 %>%
  autoplot(us_gasoline) +
  labs(title = "Forecast of U.S. Gasoline Supply for 2005",
       y = "Million Barrels per Day", x = "Year")
```
